---
title: "Unit Testing"
output: html_document
date: "2025-04-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What is Unit Testing

* Unit testing is the process of writing automated testing for snippets of code, typically function objects
<br><br>
* Each test validates whether the function behaves as expected given inputs
  * Ensure functions work 
  * Catch bugs early
  * Helps to document expected behavior
<br><br>
* Unit implies “single”, so each unit test should test each function independently

## Unit Testing in R

* `testthat` is the most built-in popular R package for writing unit tests in R
  * Python equivalents: pytest, unittest, doctest, nose2
<br><br>
* `usethis` helps to manage and set up tests in packages
  * Python does not have a single package equivalent, but combinations of tools like cookiecutter, pytest, and setuptools replicate it
<br><br>
* General workflow:
  1. Use the usethis package to initialize testing directory and files
  2. Write the tests for each function object
  3. Run tests using the devtools package

##  `usethis` and Unit Testing
* Package in R used to help automate repetitive setup tasks in R packages. Helps to setup:
  * Packages and infrastructure
  * Documentation
  * Git/Github
<br><br>
* For unit testing, helps to setup and organize the testthat framework
  * Initializing a directory within the project for tests
  * Creating test files for function objects
  * Easily run testing and get metric breakdown of successes/failures

## Getting Started

### Setting up the Package

For all projects you're working on, you should start by generating a package in R to consolidate all your project code.

```{r eval=FALSE}
## create a new package in the root folder

usethis::create_package("testingPackage")

## fill in Description and Add license

usethis::use_mit_license("testingPackage/")

## create a ReadMe file and skeleton

usethis::use_readme_md("testingPackage/")
```

### 1. Creating the Testing Directory

Unit testing packages will often use dedicated directories for the test files. Allowing for 
separation of domains (main code, modules, test code, etc.), organized testing location to allow
for easy discovery, and avoiding accidental execution. 

`usethis::use_testthat()`

* Sets up the testing folders/infrastructure for the package
* Adds testthat to Suggests attribute in DESCRIPTION file
* Creates tests/testthat/ directory and a rudimentary test file

```{r eval=FALSE}
usethis::use_testthat()
```

### Creating Data Directories

For the same benefits of creating a dedicated directory for test files, dedicating a directory for
data files has the same benefits (i.e. organization, execution avoidance).

`usethis::use_data_raw()`

* Creates data-raw/ folder
* Adds a script (data-raw/dataset_name.R) for importing and cleaning raw data

```{r eval=FALSE}
usethis::use_data_raw()
```

## 2. Create a test file

`usethis::use_test("function_name")`

* Adds a new test file in `tests/testthat/` directory for the specified function or feature
* File naming convention is to attach `test-` to the beginning of `function_name` (e.g. `test-function_name.R`)
* It is not necessary to have one .R file per function, it is often cleaner and more practical to group functions 
that are closely related to help avoid cluttering the `tests` directory. 

```{r eval=FALSE}
usethis::use_test(“function_name”)
usethis::use_test(“gamma”)
```

### Types of testing functions

The `testthat::test_that(desc, code)` function takes in as arguments:
* `desc` the name of the test
* `code` the test code containing a series of expectations being evaluated for testing, where the expectations are
defined via `expect_*()` functions

#### Common expect_*() Functions

- `expect_equal()` expects a set of outputs to be an exact value match
- `expect_true()` expects a result of logical TRUE
- `expect_false()` expects a result of logical FALSE
- `expect_snapshot()` expects an output to of code (value, table, etc.) to be the same from a previous run
- `expect_error()` expects an error to be produced, but does not crash the program
  - Uses the `stopifnot()` function to throw an error back to the `expect_error()` function call

## 3. Confirm your tests run

`devtools::test()`

* Runs all of the tests in the directory `tests/testthat/` using various `testthat` functions
* Will output all of the tests warnings/passes/failures
* Provide error messages for the warnings and failures

```{r eval=FALSE}
devtools::test()
```

`devtools::test_coverage()`

* Measures how much of the executed test code covers your code base by line
  * Excludes blank lines, comments, and Roxygen headers
* Highlights the lines of code covered and provides percentages to for each .R file of code coverage. 

```{r eval=FALSE}
devtools::test_coverage()
```